\documentclass{AnantDocumentation}

\backgroundsetup{
	contents=""
}

\lstset{
	backgroundcolor=\color{white},
	basicstyle=\ttfamily\small,
	breaklines=true,
	frame=single,
	numbers=left,
	numberstyle=\tiny,
	keywordstyle=\color{blue},
	commentstyle=\color{gray},
	stringstyle=\color{red},
	showstringspaces=false
}

\title{ECE F344: Information Theory and Coding \\ Assignment 1}
\author{Pranav Chandra N. V. \\ 2023AAPS0013P}

\begin{document}
\maketitle
\tableofcontents
\newpage

\section{Introduction}
The goals of this assignment, as per the brief, are listed below:
\begin{itemize}
	\item Implement Huffman Encoding and Decoding
	\item Implement Shannon-Type Encoding
	\item Implement Shannon Encoding
	\item Implement Shannon-Fano (Fano) Encoding
	\item Implement D-ary (non-binary) Huffman Encoding
	\item Provide a comparative statement between the different encoding efficiencies
\end{itemize}
The above goals were stated as the minimum required targets. All set targets were achieved. The submission is able to perform the following:
\begin{itemize}
	\item Accept input either via stdin (prompted string) or as a \texttt{.txt} file argument.
	\item Huffman encoding and decoding (binary and D-ary).
	\item D-ary Huffman encoding with $2 \le D \le 9$.
	\item Shannon-Type encoding and decoding.
	\item Shannon encoding and decoding.
	\item Shannon-Fano encoding and decoding.
	\item Entropy calculation and a comparison table of all average code lengths.
\end{itemize}

\section{Assumptions}
\begin{itemize}
	\item Symbol probabilities are derived from character frequencies in the input string:
	      $P(x_i) = \frac{\text{count}(x_i)}{N}$, where $N$ is the total number of characters.
	\item All whitespace and punctuation characters in the input are treated as symbols.
	\item For D-ary Huffman, if the number of symbols $n$ does not satisfy $(n-1) \bmod (D-1) = 0$,
	      dummy symbols with probability 0 are silently appended to satisfy the condition.
	\item All generated codes are prefix-free by construction, enabling unambiguous decoding.
	\item Entropy and average code lengths are computed in bits (base-2 logarithm), regardless of D.
	\item The input string is limited to 10\,000 characters.
\end{itemize}

\section{Methodology}

\subsection{Data Structures}
While implementing these algorithms on paper is generally straightforward, implementation in C required careful management of data. The primary challenge was building and traversing trees while keeping all code values associated with their original symbols. The node structure used throughout is:

\begin{lstlisting}[language=C]
typedef struct node {
    // Node Properties
    int id;
    float prob;
    char symbol;
    // Tree Creation Properties (generalised to D-ary)
    struct node* children[MAX_D];
    int child_count;
    // Huffman Code Values
    char code[100];
    int code_len;
    // Shannon-Type Code Values
    char st_code[100];
    int st_code_len;
    // Shannon Code Values
    char sh_code[100];
    int sh_code_len;
    // Fano Code Values
    char fano_code[100];
    int fano_code_len;
} node;
\end{lstlisting}

Each node stores the symbol, its probability, tree linkage (a D-ary array of child pointers), and the codes assigned by all four methods. This keeps all data portable and co-located.

\subsection{Huffman Coding}
Huffman coding constructs an optimal prefix-free code by greedily merging the $D$ least-probable symbols at each step:
\begin{enumerate}
	\item Sort all symbols by probability (ascending).
	\item Merge the $D$ lowest-probability nodes into one parent node whose probability is their sum.
	\item Repeat until one root node remains.
	\item Traverse the tree: assign digit $0, 1, \ldots, D{-}1$ to each branch, recording the path to each leaf as its codeword.
\end{enumerate}
For binary Huffman, $D=2$. For D-ary, dummy symbols with $p=0$ are added if needed so that $(n-1) \bmod (D-1) = 0$.

\subsection{Shannon-Type Coding}
Symbols are sorted by descending probability. The codeword length for symbol $x_i$ is assigned as:
\[
	l_i = \left\lceil \log_2 \frac{1}{P(x_i)} \right\rceil
\]
Actual codewords are then assigned canonically by incrementing a counter and left-shifting when the length increases, producing a valid prefix-free code.

\subsection{Shannon Coding (Shannon-Fano-Elias)}
Symbols are sorted by descending probability. The cumulative probability up to (but not including) symbol $i$ is:
\[
	F_i = \sum_{k=0}^{i-1} P(x_k)
\]
The codeword for $x_i$ is the first $l_i = \lceil \log_2(1/P(x_i)) \rceil$ bits of the binary expansion of $F_i$.

\subsection{Shannon-Fano Coding}
Symbols are sorted by descending probability. The list is recursively split into two groups whose total probabilities are as equal as possible. The top group receives a \texttt{0} prefix and the bottom group a \texttt{1} prefix at each level. This continues until each group contains a single symbol.

\subsection{Entropy and Efficiency}
The source entropy (in bits) is:
\[
	H = -\sum_{i} P(x_i) \log_2 P(x_i)
\]
The average codeword length for a given coding scheme is:
\[
	L_{\text{avg}} = \sum_{i} P(x_i) \cdot l_i
\]
Coding efficiency is defined as:
\[
	\eta = \frac{H}{L_{\text{avg}}} \times 100\%
\]

\subsection{Program Flow}
\begin{enumerate}
	\item Executable is run (with or without a \texttt{.txt} file argument).
	\item If no argument is passed, the user is prompted for a string.
	\item The input is parsed into unique symbols; $P(x_i)$ is computed from character frequency.
	\item The user selects $D$ for D-ary Huffman.
	\item All four coding schemes generate their codewords and store them back into the nodes.
	\item For each scheme, the bitstream for the full input string is generated and printed.
	\item Each bitstream is decoded and the recovered string is printed.
	\item A comparison table of average lengths and the entropy bound is displayed.
\end{enumerate}

\section{Operation}
The code utilizes no external libraries, save for the standard C libraries \lstinline[language=C]{stdio.h}, \lstinline[language=C]{string.h}, \lstinline[language=C]{stdlib.h}, and \lstinline[language=C]{math.h}.

A \texttt{Makefile} is provided. Run \lstinline{make} in the \texttt{src/} directory to compile. The \texttt{itc} executable will be produced. Usage:
\begin{itemize}
	\item \texttt{./itc} --- prompts for a string via stdin.
	\item \texttt{./itc input.txt} --- reads the contents of \texttt{input.txt}.
\end{itemize}
After entering the string, the user is prompted to select $D$ (the arity of the Huffman tree). All coding results and the comparison summary are printed to stdout.

\newpage
\section{Sample Output}
The following output was produced using the input string \texttt{aaaabcdefg}, which corresponds to the 7-symbol example from the assignment brief with approximate probabilities $p_1=0.4$ and $p_2=\cdots=p_7=0.1$.

\begin{lstlisting}
Entropy H = 2.521928 bits/symbol

=== 2-ary Huffman Code ===
Symbol | Probability | Huffman Code | Length
a      | 0.400000    | 11           | 2
b      | 0.100000    | 000          | 3
c      | 0.100000    | 001          | 3
d      | 0.100000    | 010          | 3
e      | 0.100000    | 011          | 3
f      | 0.100000    | 100          | 3
g      | 0.100000    | 101          | 3
Average Length = 2.600000 | Efficiency = 97.00%
Huffman bitstream:
11111111000001010011100101
Decoded string: aaaabcdefg

=== Shannon-Type Code ===
Symbol | Probability | Shannon-Type Code | Length
a      | 0.400000    | 00                | 2
b      | 0.100000    | 0100              | 4
c      | 0.100000    | 0101              | 4
d      | 0.100000    | 0110              | 4
e      | 0.100000    | 0111              | 4
f      | 0.100000    | 1000              | 4
g      | 0.100000    | 1001              | 4
Average Length = 3.200000 | Efficiency = 78.81%
Shannon-Type bitstream:
00000000010001010110011110001001
Shannon-Type decoded: aaaabcdefg

=== Shannon Code ===
Symbol | Probability | Shannon Code | Length
a      | 0.400000    | 00           | 2
b      | 0.100000    | 0110         | 4
c      | 0.100000    | 1000         | 4
d      | 0.100000    | 1001         | 4
e      | 0.100000    | 1011         | 4
f      | 0.100000    | 1100         | 4
g      | 0.100000    | 1110         | 4
Average Length = 3.200000 | Efficiency = 78.81%
Shannon bitstream:
00000000011010001001101111001110
Shannon decoded: aaaabcdefg

=== Shannon-Fano Code ===
Symbol | Probability | Shannon-Fano Code | Length
a      | 0.400000    | 00                | 2
b      | 0.100000    | 01                | 2
c      | 0.100000    | 100               | 3
d      | 0.100000    | 101               | 3
e      | 0.100000    | 1100              | 4
f      | 0.100000    | 1101              | 4
g      | 0.100000    | 111               | 3
Average Length = 2.700000 | Efficiency = 93.40%
Fano bitstream:
000000000110010111001101111
Fano decoded: aaaabcdefg

=== Comparison Summary ===
Coding Scheme        | Avg Length
Entropy Bound        | 2.521928
Huffman              | 2.600000
Shannon-Type         | 3.200000
Shannon              | 3.200000
Shannon-Fano         | 2.700000
\end{lstlisting}

As expected, Huffman coding achieves the closest average length to the entropy bound, followed by Shannon-Fano. Shannon and Shannon-Type both yield a longer average length for this input.

\begin{comment}
\section{Source Code}

\subsection{global\_req.h}
\lstinputlisting[language=C]{../src/global_req.h}

\subsection{global\_req.c}
\lstinputlisting[language=C]{../src/global_req.c}

\subsection{main.c}
\lstinputlisting[language=C]{../src/main.c}

\subsection{huffman.h}
\lstinputlisting[language=C]{../src/huffman.h}

\subsection{huffman.c}
\lstinputlisting[language=C]{../src/huffman.c}

\subsection{shannon\_type.h}
\lstinputlisting[language=C]{../src/shannon_type.h}

\subsection{shannon\_type.c}
\lstinputlisting[language=C]{../src/shannon_type.c}

\subsection{shannon.h}
\lstinputlisting[language=C]{../src/shannon.h}

\subsection{shannon.c}
\lstinputlisting[language=C]{../src/shannon.c}

\subsection{fano.h}
\lstinputlisting[language=C]{../src/fano.h}

\subsection{fano.c}
\lstinputlisting[language=C]{../src/fano.c}
\end{comment}


\end{document}

